---
title: "Pocillopora meta-analysis VCF SNP filtering"
author: "Mike Connelly"
date: "01/21/2022"
output: html_document
---

```{r setup, include=FALSE}
# O'Leary et al. 
# modified from https://github.com/sjoleary/SNPFILT/blob/master/filteringsnapper.Rmd
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

```
```{r}
library("tidyverse")
source("./R/VCFfilterstats_edit.R")
source("./R/VCFfilterviz.R")
# need to build custom functions in the repository (ex. multiplot())
```

## Preliminary filtered dataset stats
The raw VCF was filtered on Hydra for loci with <25% missing data, Q > 30, depth 20 - 300, and minor allele frequency > 0.01
The preliminary VCF contains 720,086 sites in 217 individuals.
Target several thousand (~1,000 - 10,000) SNPs for downstream analyses
```{bash vcftools_query_raw}
# vcf file, out directory, outfile prefix
bash ./bash/vcftools_query.sh \
./outputs/phylotrans_pdam/metapocillopora_004_filtered_primary.vcf \
./outputs/phylotrans_pdam/metapocillopora_004_primary \
metapocillopora_004_primary
```
### Query stats
```{r raw_stats}
ind_stats_primary <- read.ind.stats(dir = "./outputs/phylotrans_pdam/metapocillopora_004_primary", vcf = "metapocillopora_004_primary")
loc_stats_primary <- read.loc.stats(dir = "./outputs/phylotrans_pdam/metapocillopora_004_primary", vcf = "metapocillopora_004_primary")
site_qual_raw <- read.table("./outputs/phylotrans_pdam/metapocillopora_004_primary/metapocillopora_004_primary.lqual",
                            header = TRUE, stringsAsFactors = FALSE) %>%
  mutate(PROB = 10^(-QUAL/10))
```

### Visualize preliminary dataset stats
```{r raw_dataset_visualization}
vcfqvis("metapocillopora_004_primary", ind_stats_primary, loc_stats_primary)
```
```{r raw_dataset_missing-data}
# Visualize missing data among individuals and loci
pdf("./outputs/phylotrans_pdam/filtered_vcf_viz/meta_004_primary.pdf", height = 8.5, width = 11)
vcf_raw_viz <- vcfqvis_miss("metapocillopora_004_primary", ind_stats_primary, loc_stats_primary, title = "Primary VCF")
dev.off()
```

## Filtering strategy 
* Step 1: Remove loci with genotype call rate < 25% (missing data > 75%)
* Step 2: Remove individuals with missing data > 95%
* Step 3: Remove loci with depth < 20 
* Step 4: Remove loci with minor allele frequency < 5%

### Step 1: Remove loci with genotype call rate < 25%
```{bash vcftools_filter_1}
# Step 1: Remove loci with genotype call rate < 25%
bash ./bash/vcftools_filter_maxmissing.sh \
./outputs/phylotrans_pdam/metapocillopora_004_filtered_primary.vcf \
./outputs/phylotrans_pdam/filtered_vcfs/max-missing-0.75 \
0.75
```
The filtered VCF contains 335,796 sites in 217 individuals (46.6% of total sites)

```{bash vcftools_query_1}
# vcf file, out directory, outfile prefix
bash ./bash/vcftools_query.sh \
./outputs/phylotrans_pdam/filtered_vcfs/max-missing-0.75.recode.vcf \
./outputs/phylotrans_pdam/max-missing-0.75 \
max-missing-0.75
```
```{r read vcf query stats and viz}
#
ind_stats_1 <- read.ind.stats(dir = "./outputs/phylotrans_pdam/max-missing-0.75", vcf = "max-missing-0.75")
loc_stats_1 <- read.loc.stats(dir = "./outputs/phylotrans_pdam/max-missing-0.75", vcf = "max-missing-0.75")
#
pdf("./outputs/phylotrans_pdam/filtered_vcf_viz/max-missing-0.75_vcf.pdf", height = 8.5, width = 11)
vcfqvis_miss(vcf = "max-missing-0.75", ind_stats_1, loc_stats_1, title = " 25% max-missing VCF")
dev.off()
```
```{r raw_dataset_missing-data}
# Visualize missing data among individuals and loci
pdf("./outputs/phylotrans_pdam/filtered_vcf_viz/meta_step1_vcf.pdf", height = 8.5, width = 11)
vcf_raw_viz <- vcfqvis_miss("max-missing-0.75", ind_stats_1, loc_stats_1, title = "Step 1: Max-missing-0.25")
dev.off()
```

### Step 2: Remove individuals with missing data > 75%
```{r}
imissing <- read.table("./outputs/phylotrans_pdam/max-missing-0.75/max-missing-0.75.imiss",
                       header = TRUE, stringsAsFactors = FALSE)
# plot
ggplot(imissing, aes(x = F_MISS)) +
  geom_histogram(binwidth = .01, color = "black", fill = "grey95") +
  geom_vline(aes(xintercept = 0.85),
                 color = "red", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = mean(F_MISS, na.rm = TRUE)),
                 color = "black", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = 0.25),
                 color = "blue", linetype = "dashed", size = 1) +
  scale_x_continuous(limits = c(0, 1)) +
  labs(x = "missing data per indv") 
```
```{r}
#
indv_missing_0.75 <- imissing %>%
  dplyr::filter(F_MISS > 0.25) %>%
  select(INDV)
#
write.table(indv_missing_0.75, "./outputs/phylotrans_pdam/max-missing-0.75/indv_missing_0.75.txt",
            col.names = FALSE, row.names = FALSE, quote = FALSE)
```
```{bash vcftools_filter_2}
# Step 2: Remove individuals with missing data > 95%
bash ./bash/vcftools_filter_remove.sh \
./outputs/phylotrans_pdam/filtered_vcfs/max-missing-0.75.recode.vcf \
./outputs/phylotrans_pdam/filtered_vcfs/max-missing-0.75_indv-missing-0.75 \
./outputs/phylotrans_pdam/max-missing-0.75/indv_missing_0.75.txt
```
The filtered VCF contains 184,904 sites in 210 individuals (7 individuals were removed)
"ConnellyEAPSI_SRR10395136_1" "ConnellyEAPSI_SRR10395145_1" 
"Grupstra2022_SRR16897910"    "Grupstra2022_SRR16897911"   
"Grupstra2022_SRR16897912"    "Grupstra2022_SRR16897913"    
"Grupstra2022_SRR16897914"   

```{bash vcftools_query_2}
# vcf file, out directory, outfile prefix
bash ./bash/vcftools_query.sh \
./outputs/phylotrans_pdam/filtered_vcfs/max-missing-0.75_indv-missing-0.75.recode.vcf \
./outputs/phylotrans_pdam/max-missing-0.75_indv-missing-0.75/ \
max-missing-0.75_indv-missing-0.75
```
```{r read vcf query stats and viz 2}
#
ind_stats_2 <- read.ind.stats(dir = "./outputs/phylotrans_pdam/max-missing-0.75_indv-missing-0.75", vcf = "max-missing-0.75_indv-missing-0.75")
loc_stats_2 <- read.loc.stats(dir = "./outputs/phylotrans_pdam/max-missing-0.75_indv-missing-0.75", vcf = "max-missing-0.75_indv-missing-0.75")
#
vcfqvis(vcf = "max-missing-0.75_indv-missing-0.75.recode", ind_stats_2, loc_stats_2)
```
```{r}
# Visualize missing data among individuals and loci
pdf("./outputs/phylotrans_pdam/filtered_vcf_viz/merge_step2_vcf.pdf", height = 8.5, width = 11)
vcf_raw_viz <- vcfqvis_miss("max-missing-0.75_indv-missing-0.75", ind_stats_2, loc_stats_2, title = "Step 2: Max-missing-0.25, Indv-missing-0.75")
dev.off()
```

### Step 3: Remove loci with genotype call rate < 25%
```{bash vcftools_filter_1}
# Step 1: Remove loci with genotype call rate < 10%
bash ./bash/vcftools_filter_maxmissing.sh \
./outputs/phylotrans_pdam/filtered_vcfs/max-missing-0.75_indv-missing-0.75.recode.vcf \
./outputs/phylotrans_pdam/filtered_vcfs/max-missing-0.75_indv-missing-0.75_max-missing_0.1 \
0.90
```

The filtered VCF contains 2,714 sites in 61 individuals.
```{bash vcftools_query_1}
# vcf file, out directory, outfile prefix
bash ./bash/vcftools_query.sh \
./outputs/phylotrans_pdam/filtered_vcfs/max-missing-0.75_indv-missing-0.75_max-missing_0.1.recode.vcf \
./outputs/phylotrans_pdam/max-missing-0.75_indv-missing-0.75_max-missing_0.1/ \
max-missing-0.75_indv-missing-0.75_max-missing_0.1
```
```{r read vcf query stats and viz}
#
ind_stats_3 <- read.ind.stats(dir = "./outputs/phylotrans_pdam/max-missing-0.75_indv-missing-0.75_max-missing_0.1", vcf = "max-missing-0.75_indv-missing-0.75_max-missing_0.1")
loc_stats_3 <- read.loc.stats(dir = "./outputs/phylotrans_pdam/max-missing-0.75_indv-missing-0.75_max-missing_0.1", vcf = "max-missing-0.75_indv-missing-0.75_max-missing_0.1")
#
vcfqvis(vcf = "max-missing-0.75_indv-missing-0.75_max-missing_0.1", ind_stats_3, loc_stats_3)
```
```{r}
# Visualize missing data among individuals and loci
pdf("./outputs/phylotrans_pdam/filtered_vcf_viz/merge_step3_vcf.pdf", height = 8.5, width = 11)
vcf_raw_viz <- vcfqvis_miss("max-missing-0.05_indv-missing-0.85_depth-20", ind_stats_3, loc_stats_3, title = "Step 3: Max-missing-0.05, Indv-missing-0.85, Depth-20")
dev.off()
```

### Step 4: Remove loci with minor allele frequency < 5%
```{bash vcftools_filter_1}
# Step 4: Remove loci with minor allele frequency < 5%
bash ./bash/vcftools_filter_maf.sh \
./outputs/phylotrans_pdam/filtered_vcfs/max-missing-0.05_indv-missing-0.85_depth-20.recode.vcf \
./outputs/phylotrans_pdam/filtered_vcfs/max-missing-0.05_indv-missing-0.85_depth-20_maf-0.05 \
 0.05
```
The filtered VCF contains 1,417 sites in 61 individuals.
```{bash vcftools_query_1}
# vcf file, out directory, outfile prefix
bash ./bash/vcftools_query.sh \
./outputs/phylotrans_pdam/filtered_vcfs/max-missing-0.05_indv-missing-0.85_depth-20_maf-0.05.recode.vcf \
./outputs/phylotrans_pdam/max-missing-0.05_indv-missing-0.85_depth-20_maf-0.05 \
max-missing-0.05_indv-missing-0.85_depth-20_maf-0.05
```
```{r read vcf query stats and viz}
#
ind_stats_4 <- read.ind.stats(dir = "./outputs/phylotrans_pdam/max-missing-0.05_indv-missing-0.85_depth-20_maf-0.05", vcf = "max-missing-0.05_indv-missing-0.85_depth-20_maf-0.05")
loc_stats_4 <- read.loc.stats(dir = "./outputs/phylotrans_pdam/max-missing-0.05_indv-missing-0.85_depth-20_maf-0.05", vcf = "max-missing-0.05_indv-missing-0.85_depth-20_maf-0.05")
#
vcfqvis(vcf = "max-missing-0.05_indv-missing-0.85_depth-20_maf-0.05", ind_stats_4, loc_stats_4)
```
```{r}
# Visualize missing data among individuals and loci
pdf("./outputs/phylotrans_pdam/filtered_vcf_viz/merge_step4_vcf.pdf", height = 8.5, width = 11)
vcf_raw_viz <- vcfqvis_miss("max-missing-0.05_indv-missing-0.85_depth-20_maf-0.05", ind_stats_4, loc_stats_4, title = "Step 4: Max-missing-0.05, Indv-missing-0.85, Depth-20, MAF-0.05")
dev.off()
```

## Filtering strategy #2
* Step 1: Remove loci with genotype call rate < 5% (missing data > 95%)
* Step 2: Remove individuals with missing data > 95%
* Step 3: Remove loci with depth < (10 - 25)
* Step 4: Remove loci with minor allele frequency < 5%
* Step 5: Remove loci that are not in Hardy-Weinberg equilibrium

## Custom filtering strategy
### Step 1: Identify individuals with > 90% missing data:
```{r, fig.height=4, fig.width=5}
imissing <- read.table("./outputs/phylotrans_pdam/metapocillopora_001.imiss",
                       header = TRUE, stringsAsFactors = FALSE)
# plot
ggplot(imissing, aes(x = F_MISS)) +
  geom_histogram(binwidth = .01, color = "black", fill = "grey95") +
    geom_vline(aes(xintercept = 0.95),
                 color = "red", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = mean(F_MISS, na.rm = TRUE)),
                 color = "black", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = 0.25),
                 color = "blue", linetype = "dashed", size = 1) +
  scale_x_continuous(limits = c(0, 1)) +
  labs(x = "missing data per indv") 

indv_missing_0.95 <- imissing %>%
  dplyr::filter(F_MISS > 0.95) %>%
  select(INDV)

# write.table(indv_missing_0.95, "./outputs/phylotrans_pdam/indv_missing_95",
            # col.names = FALSE, row.names = FALSE, quote = FALSE)
```
```{r}
imissing <- read.table("./outputs/phylotrans_pdam/max-missing-0.05/etp_gbs_max-missing-0.05.imiss",
                       header = TRUE, stringsAsFactors = FALSE)
# plot
ggplot(imissing, aes(x = F_MISS)) +
  geom_histogram(binwidth = .01, color = "black", fill = "grey95") +
  geom_vline(aes(xintercept = 0.95),
                 color = "red", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = mean(F_MISS, na.rm = TRUE)),
                 color = "black", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = 0.25),
                 color = "blue", linetype = "dashed", size = 1) +
  scale_x_continuous(limits = c(0, 1)) +
  labs(x = "missing data per indv") 

indv_missing_0.95 <- imissing %>%
  dplyr::filter(F_MISS > 0.95) %>%
  select(INDV)

write.table(indv_missing_0.95, "./outputs/phylotrans_pdam/max-missing-0.05/indv_missing_0.95",
            col.names = FALSE, row.names = FALSE, quote = FALSE)
```

```{r}
imissing <- read.table("./outputs/phylotrans_pdam/max-missing-0.25/etp_gbs_max-missing-0.25.imiss",
                       header = TRUE, stringsAsFactors = FALSE)

ggplot(imissing, aes(x = F_MISS)) +
  geom_histogram(binwidth = .01, color = "black", fill = "grey95") +
  geom_vline(aes(xintercept = 0.95),
                 color = "red", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = mean(F_MISS, na.rm = TRUE)),
                 color = "black", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = 0.25),
                 color = "blue", linetype = "dashed", size = 1) +
  scale_x_continuous(limits = c(0, 1)) +
  labs(x = "missing data per indv") 

indv_missing_0.95 <- imissing %>%
  dplyr::filter(F_MISS > 0.95) %>%
  select(INDV)

write.table(indv_missing_0.95, "./outputs/phylotrans_pdam/max-missing-0.25/indv_missing_0.95",
            col.names = FALSE, row.names = FALSE, quote = FALSE)
```

```{r}
ind_stats_A <- read.ind.stats(dir = "~/Desktop/vcfquery/max-missing-0.05_indv-missing-0.95", vcf = "etp_gbs_max-missing-0.05_indv-missing-0.95")
loc_stats_A <- read.loc.stats(dir = "~/Desktop/vcfquery/max-missing-0.05_indv-missing-0.95", vcf = "etp_gbs_max-missing-0.05_indv-missing-0.95")
```
```{r}
pdf("~/Desktop/vcfqvis_test.pdf", height = 11, width = 8.5)
vcfqvis(vcf = "etp_gbs_max-missing-0.05_indv-missing-0.95", ind_stats_A, loc_stats_A)
dev.off()
```
```{r}
ind_stats_B <- read.ind.stats(dir = "~/Desktop/vcfquery/max-missing-0.25_indv-missing-0.95", vcf = "etp_gbs_max-missing-0.25_indv-missing-0.95")
loc_stats_B <- read.loc.stats(dir = "~/Desktop/vcfquery/max-missing-0.25_indv-missing-0.95", vcf = "etp_gbs_max-missing-0.25_indv-missing-0.95")
```
```{r}
pdf("~/Desktop/vcfqvis_test.pdf", height = 11, width = 8.5)
vcfqvis(vcf = "etp_gbs_max-missing-0.25_indv-missing-0.95", ind_stats_B, loc_stats_B)
dev.off()
```
```{r}
ind_stats_C <- read.ind.stats(dir = "~/Desktop/vcfquery/max-missing-0.25_indv-missing-0.95_maf_0.01_hwe-0.05", vcf = "max-missing-0.25_indv-missing-0.95_maf_0.01_hwe-0.05")
loc_stats_C <- read.loc.stats(dir = "~/Desktop/vcfquery/max-missing-0.25_indv-missing-0.95_maf_0.01_hwe-0.05", vcf = "max-missing-0.25_indv-missing-0.95_maf_0.01_hwe-0.05")
```
```{r}
pdf("~/Desktop/vcfqvis_test_2.pdf", height = 11, width = 8.5)
vcfqvis(vcf = "max-missing-0.25_indv-missing-0.95_maf_0.01_hwe-0.05", ind_stats_C, loc_stats_C)
dev.off()
```

