---
title: "Pocillopora_meta-transcriptome_analysis"
author: "Mike Connelly"
date: "1/11/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, cache = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
options(stringsAsFactors = FALSE)
```

## Setup packages and working directories
```{r packages, message=FALSE, include=FALSE}
# Include all packages needed for the entire transcriptome analysis and downstream visualizations
library("tidyverse")
# Essential RNAseq analysis packages
library("DESeq2")
library("PCAtools")
library("apeglm")
library("WGCNA")
library("dynamicTreeCut")
library("flashClust")
# library("genefilter")
library("variancePartition")
library("doParallel")
# library("limma")
# library("adegenet")
# GO/KOG enrichment packages and other functions
# library("KOGMWU")
# source("./R/GO_MWU/gomwu.functions.R")
# Data visualization packages
# library("pheatmap")
# library("ComplexHeatmap")
# library("VennDiagram")
# library("eulerr")
# library("UpSetR")
# Graphics packages
# library("cowplot")
# library("patchwork")
# library("gridExtra")
# library("ggthemes")
# library("ggpubr")
# library("ggrepel")
# library("ggnewscale")
# library("RColorBrewer")
# library("circlize")
# library("stringr")
# library("extrafont")
# library("extrafontdb")
# extended visualization functions
source("./R/pocillopora_meta_functions.R")
```
```{r}
theme_set(theme_bw())
```

## Import sample metadata
```{r factor levels}
study_levels <- c("Mayfield2014", "Vidal-Dupiol2014", "Yuan2017", "Zhou2017", "Brener-Raffali2018", "Delgadillo-Nuno2018", "Tang2018", "Wecker2018", "Zhou2018", "Poquita-Du2019", "Zhou2019", "Buitrago-Lopez2020", "Chuang-Mitarai2020", "Li2020", "ConnellyEAPSI", "Grupstra2022")
```
```{r sample_data, echo=TRUE}
# Import sample metadata 
samples <- read_csv("data/SampleData_MetaPocillopora_012722.csv")
```
```{r coldata}
# Get sample data into tibble
coldata <- samples %>%
  dplyr::filter(Include_Counts == TRUE) %>%
  # dplyr::filter(Treatment == "control") %>% 
  as_tibble(.) %>%
  dplyr::arrange(desc("Sample")) %>% #Order rows                    
  column_to_rownames(var = "Sample")  
```

## Import *Pocillopora* coral counts data and gene annotation
```{r import and merge featureCounts}
#get files names
fileList = list.files(path = "./outputs/featureCounts_pdam/", pattern = "*counts", full.names = TRUE)
print('Merging the following feature counts files:')
print(paste(length(fileList), 'total'))
for (f in fileList){
  print(f)
}
#read in function
read_dat = function(x){
  print(paste(x,'...',sep=''))
  read.table(x, sep="\t", header = TRUE) %>% 
  as_tibble()
}
#read in and join the datasets
print('Reading in and joining...')
dat <- fileList %>%
  purrr::map(function(x) read_dat(x)) %>% 
  purrr::reduce(full_join, by = c('Geneid', 'Chr', 'Start', 'End', 'Strand', 'Length'))
print('Done.')
#print summary
nGenes = nrow(dat)
nSample = ncol(dat)-6
print(paste(nGenes, 'total genes'))
print(paste(nSample, 'total samples'))
```
```{r coral_countdata}
# Import and tidy counts data from each study
countdata <- dat
# Set Gene ID's as row names
countdata <- countdata %>% column_to_rownames("Geneid")
# row.names(countdata) <- countdata$Geneid
# Remove first six columns (Geneid, chr, start, end, strand, length)
countdata <- countdata[ ,6:ncol(countdata)]
# Remove file prefixes and suffixes to leave SRR accession number
  colnames(countdata) <- gsub("X.scratch.nmnh_corals.connellym.projects.MetaPocillopora.outputs.STARalign_pdam.", "", colnames(countdata))
  colnames(countdata) <- gsub("_pdam_Aligned.out.bam$", "", colnames(countdata))
  colnames(countdata) <- gsub("_1$", "", colnames(countdata))
  colnames(countdata) <- gsub("^.*\\..*_", "", colnames(countdata))
  # colnames(countdata) <- gsub("\\.", "-", colnames(countdata))
```
```{r countdata_tidy}
# Select counts for samples present in coldata
countdata_select <- countdata[, colnames(countdata) %in% rownames(coldata)]

# Sort countdata by P. damicornis gene ID
countdata_sorted_coral <- as.data.frame(countdata_select[sort(rownames(countdata_select)), order(match(colnames(countdata_select),row.names(coldata)))]) 

# Sanity check
all(colnames(countdata_sorted_coral) == rownames(coldata))
```
```{r coral_gene_annotation}
gene_annotation_coral <- read.delim(file = "./data/pdam_genome_annotations.tsv", header = T) %>% arrange(ID)
rownames(gene_annotation_coral) <- gene_annotation_coral$ID
# Check gene feature annotation and countdata rowname order coherence
all(rownames(countdata_sorted_coral) == gene_annotation_coral$ID)
all(rownames(countdata_sorted_coral) ==  rownames(gene_annotation_coral))
# Obtain KOG annotations for P. damicornis genome
gene2kog_coral <- gene_annotation_coral %>%
  dplyr::select(ID, KOG_Class) %>% 
  filter(KOG_Class != "")
```
```{r}
gene_annotation <- gene_annotation_coral
```

## Create DESeq datasets
### *Pocillopora* coral host
```{r create_dds_coral}
# Create full DESeqDataSet
dds_coral <- DESeqDataSetFromMatrix(countData = countdata_sorted_coral,
                              colData = coldata,
                              design = ~1)
# Check annotation and dds_coral object rowname order coherence
all(rownames(dds_coral) == rownames(gene_annotation_coral))
# Add gene feature annotation to DESeqDataSets
mcols(dds_coral) <- cbind(mcols(dds_coral), gene_annotation_coral)
# Subset DESeqDataSet
# Remove genes with counts less than 10 in 90% of samples
keep <- rowSums(counts(dds_coral) >= 10) > (ncol(countdata_sorted_coral)*0.90)
dds_coral <- dds_coral[keep, ]
# Normalize expression data for visualization purposes using VST tranformation
vsd_coral <- vst(dds_coral, blind = TRUE) # use blind = TRUE to not account for experimental design
```

## Visualize global gene expression
```{r colors}
gg_color_hue <- function(n) {
  hues = seq(15, 375, length = n + 1)
  hcl(h = hues, l = 65, c = 100)[1:n]
}
```
```{r choose_vsd, include=FALSE}
# Choose vsd for host/symbiont analysis with PCoA and WGCNA
vsd <- vsd_coral
# vsd <- vsd_symbiont
```
## Figure 3: PCA of overall gene expression
```{r prepare_pca_dataframe}
pca_df <- plotPCA.custom(vsd, intgroup = c("Study"), ntop = 500, returnData = TRUE,  pcs = c(1,2))
# 
pca_df<- pca_df %>% dplyr::rename("Sample" = "name") %>%
  # mutate(Sample_ID = str_replace(`Sample_ID`, "_1$", "")) %>% 
  inner_join(samples, by = "Sample") %>% 
  dplyr::rename("Study" = Study.x)
pca_df$Study <- factor(pca_df$Study, levels = study_levels, ordered = TRUE)
```
```{r gene expression pca figure}
pdf("./outputs/figures/Fig3_gene_expression_pca.pdf", height = 7.5, width = 9)
pca_title <- paste("PCA of overall gene expression (12,826 genes)", sep = "")
#
PCAtmtpercentVar <- round(100 * attr(pca_df, "percentVar"), 2)
pc1lab <- paste0( "PC", 1, ": ", PCAtmtpercentVar[1], "%")
pc2lab <- paste0( "PC", 2, ": ", PCAtmtpercentVar[2], "%")
#
pca_df %>%
  ggplot(., aes(PC1, PC2)) +
  geom_point( aes(fill = Study, shape = Species), size = 6, alpha = 0.8, stroke = 0.5, color = "black") + 
  # geom_text_repel(aes(label = Sample_ID), size = 2, max.overlaps = 100)
  scale_fill_manual(values = gg_color_hue(n_distinct(coldata$Study)), 
                    guide = guide_legend(override.aes = list(size = 6, shape = 21, fill = gg_color_hue(n_distinct(coldata$Study))), nrow = 8, title.position = "top")) +
  scale_shape_manual(values = c(21, 24, 22, 23),
                     guide = guide_legend(override.aes = list(size = 6, alpha = 0.8, stroke = 0.5, fill = "black"), nrow = 4, title.position = "top")) +
  theme_bw() +
  coord_fixed(12.32/21.52) + 
  theme(plot.title = element_text(size = 28),
        axis.title = element_text(size = 24),
        axis.text = element_blank(),
        legend.title = element_text(size = 24),
        legend.text = element_text(size = 20),
        legend.position = "bottom",
        legend.spacing.y = unit(0.15, "in")) +
  xlab(pc1lab) + ylab(pc2lab) +
  ggtitle(pca_title)
dev.off()
```

### Visualize global gene expression using PCAtools
```{r}
p <- pca(assay(vsd), metadata = coldata, removeVar = 0.2)
screeplot(p, axisLabSize = 18, titleLabSize = 22)
```
```{r}
# pdf("./outputs/figures/pca_biplot_temp.pdf", height = 9, width = 9)
  biplot(p, showLoadings = FALSE, 
         colby = 'Genotype',
         shape = 'Species',
    labSize = 1, pointSize = 2, sizeLoadingsNames = 1,
    legendPosition = "right")
  # dev.off()
?biplot(

)
```
```{r}
pdf("./outputs/figures/pca_pairsplot_all.pdf", height = 9, width = 9)
pairsplot(p, colby = 'Study')
dev.off
```
```{r}
eigencorplot(p,
    metavars = c('Study', 'Species', 'Treatment', 'Temperature'))
```

## VariancePartition
```{r variance_partition}
# Partition total transcriptome variance among explanatory factors including sequencing round, coral genotype, treatment, and their interaction
formula <- ~ (1|Study) + (1|Species) + (1|Genotype) + (1|Treatment)
#
cores<-detectCores()
cl <- makeCluster(cores[1]-1); registerDoParallel(cl)
varpart_total <- fitExtractVarPartModel(assay(vsd), formula, coldata)
stopCluster(cl)
# sort variables (i.e. columns) by median fraction of variance explained
vpt <- sortCols(varpart_total, FUN = mean)
```
```{r }
# Violin plot of contribution of each variable to total variance
varcols <- rev(c('#ffffcc','#a1dab4','#41b6c4','#2c7fb8','#253494'))
varplot <- plotVarPart(vpt, col = varcols)
vartitle <- expression(paste(italic("Pocillopora"), " expression variance partition"))
varplot <- varplot +
  theme(plot.title = element_text(size = 24, hjust = -0.1),
        axis.title = element_text(size = 20),
        axis.text = element_text(size = 18),
        axis.text.x = element_text(angle = 330, hjust= 0, size = 20),
        legend.title = element_text(size = 20),
        legend.text = element_text(size = 18),
        legend.spacing.y = unit(0.15, "in")) +
  ggtitle(vartitle)
print(varplot)
ggsave(varplot, filename = "./outputs/figures/variancePartition_total.pdf", width = 10, height = 5, units = "in", device = "pdf")
```
```{r varpart_sum}
vptdf <- as.data.frame(vpt) %>% rownames_to_column("gene")
varpct <- vptdf %>% pivot_longer(cols = -gene, names_to = "source", values_to = "value") %>% 
  mutate(relvar = value / sum(value)) %>%
  mutate(source = str_replace(source, "_", " ")) %>% 
  group_by(source) %>% 
  # filter(source != "Residuals") %>%
  summarise(total_pctvar_explained = sum(relvar))
```
### Supplementary Figure
```{r}
varpct %>% 
  mutate(source = fct_reorder(source, order(colnames(vpt)))) %>% 
  ggplot(aes(source, total_pctvar_explained)) +
  geom_col(aes(fill = source), color = "black") +
  ylab("Total percent variance explained") +
  scale_fill_viridis(discrete = T) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 330, hjust = 0))
```
```{r total_variance figure: powerpoint}
pdf("./outputs/figures/Fig4_variance_partition_totals.pdf", height = 6.5, width = 3.5)
var_title <- expression(paste("Variance factors"))
# 
varpct %>% 
  dplyr::mutate(source = fct_reorder(source, order(colnames(vpt)))) %>%
  # dplyr::filter(source != "Residuals") %>% 
  ggplot(aes(source, total_pctvar_explained)) +
  geom_col(aes(fill = source), color = "black", show.legend = FALSE) +
  ylab("Total percent variance") +
  scale_fill_manual(values = varcols, name = "Factor") +
  scale_y_continuous(limits = c(0, 0.5), labels = scales::percent) +
  theme_bw() +
  theme(plot.title = element_text(size = 28, hjust = 1),
        axis.title = element_text(size = 24),
        axis.title.x = element_blank(),
        axis.text = element_text(size = 24),
        axis.text.x = element_text(angle = 270, hjust = 0)) + 
  ggtitle(var_title)
dev.off()
```

## Weighted gene co-expression network analysis (WGCNA)
```{r wgcna_setup_datExpr}
allowWGCNAThreads(nThreads = 4)
# obtain vst-transformed counts
countdata_vst <- assay(vsd)
# Convert to matrix and transpose, check dimensions
datExpr <- t(countdata_vst)
dim(datExpr)
```
```{r soft_threshold}
# Find correlation power R^N that satisfies scale free critereon (SFT.R.sq>0.9)
sft <- pickSoftThreshold(datExpr, verbose=5)
sft$powerEstimate
```
### Step-by-step network construction
```{r adjacency_TOM}
# This follows the tutorial: 
# https://horvath.genetics.ucla.edu/html/CoexpressionNetwork/Rpackages/WGCNA/Tutorials/FemaleLiver-02-networkConstr-man.pdf
# 
# I have chosen the following network construction parameters for the following reasons:
# First, following the recommendations of the WGCNA developers (https://horvath.genetics.ucla.edu/html/CoexpressionNetwork/Rpackages/WGCNA/faq.html), a signed network was chosen to be able to detect positive and negative gene correlations, and the biweight midcorrelation was used since it is more robust to outliers. 

adjacency <- adjacency(datExpr,
      # Network construction arguments:  correlation, adjacency function,  and topological overlap map options
                       corFnc = "bicor", # bimidweight correlation
                       power = sft$powerEstimate, # 20 for coral
                       type = "signed") # signed
# 
TOM <- TOMsimilarity(adjacency,
                     TOMType = "signed",
                     verbose = 5)
dissTOM <- 1-TOM
# 
rm(adjacency) # may need to delete adjacency, TOM to clear up vector memory
```
```{r gene_dendrogram}
geneTree <- flashClust(as.dist(dissTOM), method = "average")
plot(geneTree, labels = FALSE, xlab="", sub="", main = "Gene clustering on TOM-based dissimilarity")
```
```{r dynamic_treecut}
# minModuleSize = 100 because I prefer large modules that lend themselves to enrichment tests with GO_MWU for easier interpretation.
minModuleSize <- 50
# Module identification using dynamic tree cut, with cutHeight = 0.99 and deepSplit = 2 based on the recommendations of the WGCNA developers
dynamicMods <- cutreeDynamic(dendro = geneTree,
                             distM = dissTOM,
                             cutHeight = 0.99,
                             deepSplit = 2,
                             pamRespectsDendro = FALSE,
                             minClusterSize = minModuleSize)
table(dynamicMods)
# Convert numeric lables into colors
dynamicColors <- labels2colors(dynamicMods, colorSeq = standardColors())
```
```{r module_dendrogram}
# Calculate eigengenes
MEList <- moduleEigengenes(datExpr, colors = dynamicColors)
MEs <- MEList$eigengenes
# Calculate dissimilarity of module eigengenes
MEDiss <- 1-cor(MEs);
# Cluster module eigengenes
METree <- flashClust(as.dist(MEDiss), method = "average");
# Plot the result
plot(METree, main = "Clustering of module eigengenes",
xlab = "", sub = "")
MEDissThres = 0.3
# Plot the cut line into the dendrogram
abline(h=MEDissThres, col = "red")
```
```{r merge_modules}
# Call an automatic merging function
# merge cutHeight = 0.3 because I want to group together modules with >70% similar module eigengene expression
mergedMods <- mergeCloseModules(datExpr, dynamicColors, cutHeight = 0.3, verbose = 5)
# The merged module colors
mergedColors <- mergedMods$colors
table(mergedColors) 
```
```{r dendrogram_modules}
pdf(file = "./outputs/figures/WGCNA_dendrogram_coral.pdf", width = 8, height = 6)
# pdf(file = "./manuscript_figures/FigS##_WGCNA_dendrogram.pdf", width = 8, height = 6)
# Plot the dendrogram and the module colors underneath
plotDendroAndColors(dendro = geneTree, 
                    colors = cbind(dynamicColors, mergedColors),
                    groupLabels = c("Dynamic modules", "Merged dynamic"),
                    dendroLabels = FALSE, hang = 0.03,
                    addGuide = TRUE, guideHang = 0.05,
                    abHeight = c(0.99))
```
```{r merged_eigengenes}
MEList <- moduleEigengenes(datExpr, colors = mergedColors)
mergedMEs <- MEList$eigengenes
```
```{r choose_modules}
moduleColors <- mergedColors#netColors#
MEs1 <- mergedMEs#netMEs#
# MEs1 <- MEs1 %>% select(-MEgrey)
MEmodule_colors <- colnames(MEs1)
module_colors <- gsub("ME", "", colnames(MEs1))
colnames(MEs1) <- module_colors
```
```{r ME_clustering}
datME <- MEs1
dissimME <- (1-t(cor(datME, method="p")))/2
hclustME <- flashClust(as.dist(dissimME), method="average" )
# Plot the eigengene dendrogram
par(mfrow=c(1,1))
plot(hclustME, main="Clustering tree based on the module eigengenes")
```
```{r uniq_modules}
# Extract all unique modules
uniqModules <- unique(colnames(MEs1))
# set in order according to ME clustering dendogram
hclustME$order
uniqModules <- uniqModules[hclustME$order]
#
uniqModules
# create ordered factor for downstream analysis
modules_factor <- factor(levels = uniqModules[hclustME$order], ordered = TRUE)
```
### Module genes and hub genes
```{r modules_genes}
# Generate GeneIDs
Genes <- colnames(datExpr)
# Output genes and annotations for each module
for (module in uniqModules)
{
# Select module genes
inModule <- (moduleColors == module)
# Get gene IDs
modGeneIDs <- Genes[inModule] #this is the correct set of gene IDs!
# Write gene annotations into a file
fileName = paste("./outputs/WGCNA_results/notes/", module, ".csv", sep="");
module_annotation <- gene_annotation[modGeneIDs, ]
write_csv(module_annotation, path = fileName)
}
```
```{r hub_genes}
hubs <- chooseTopHubInEachModule(datExpr,
                         mergedColors,
                         corFnc = "bicor", # bimidweight correlation
                         power=sft$powerEstimate, # 10
                         # power = 10,
                         type = "signed")
hub_genes <- gene_annotation[hubs, ]
hub_genes$module <- names(hubs)
row.names(hub_genes) <- NULL
hub_genes <- hub_genes %>% dplyr::select(module, everything())
```
```{r sig_hubs}
hub_genes <- hub_genes %>%
  # filter(module %in% sigModules) %>%
  dplyr::select(module, ID, Gene_Info) %>% 
  dplyr::rename("hub gene" = ID, "UniProt best match" = Gene_Info) 
```